Assessing the potential of Jacobian-free Newton-Krylov methods for cell-centred finite volume solid mechanics

### 2025-01-26 article.tex
Update authors
Update cantilever

\author*[1,2,3]{\fnm{Philip} \sur{Cardiff}}\email{philip.cardiff@ucd.ie}
\author[4]{\fnm{Ivan} \sur{Batisti\'{c}}}
%\author[5]{\fnm{Hiroaki} \sur{Nishikawa}}
\author[4]{\fnm{\v{Z}eljko} \sur{Tukovi\'{c}}}

@@

%\hl{5 - oversolving}
%Comment on over-solving => also applies to the segregated system.
%This could be a parameter we look at.
A final comment on the Jacobian-free Newton-Krylov solution algorithm is the potential importance of \emph{oversolving}.
Here, oversolving refers to solving the linear system to too tight a tolerance during the early Newton iterations, essentially wasting time when the solution is far from the true solution.
In addition, some authors %\cite{See164_and_176_in_Knoll2004}
\cite{Knoll2004} have shown Newton convergence to be worse when the earlier iterations are solved to too tight a tolerance.
The concept of oversolving also applies to segregated solution procedures and has been well-known since the early work of Demird\v{z}i\'{c} and co-workers \cite{Demirdzic1995}, where the residuals are typically reduced by one order of magnitude in the inner linear system.
%The optimal choice of residual reduction for a Jacobian-free Newton-Krylov finite volume solid mechanics procedure is explored in Section \ref{sec:test_cases}. \hl{Check: do we examine this?}
In the current work, the residual is reduced by one order of magnitude in the segregated approach and three orders in the Jacobian-free Newton-Krylov approach.

@@

Figure \ref{fig:hronTurek-results} shows the velocity field in the fluid region and displacement magnitude in the solid region at $t = 4.42$ s, corresponding to a peak in the vertical displacement of the plate free-end oscillation.
The fluid is seen to accelerate as it passes the cylinder, with regular vortices being thrown off the plate, causing it to oscillate.
\begin{figure}[htbp]
   \centering
	   \includegraphics[width=\textwidth]{figures/hronTurek-results} 
%	\subfigure[Velocity field in the fluid region and displacement magnitude in the solid region at $t = 4.42$ s]
%	{
%	   \includegraphics[width=0.8\textwidth]{figures/hronTurek-results} 
%   	}
%	\subfigure[Vertical displacement of the end of the plate vs time]
%	{
%   		\includegraphics[width=0.3\textwidth]{figures/placeholder}  
%   	}
   \caption{Elastic plate behind a rigid cylinder case results at $t = 20$ s for the mesh with $20\,032$ cells in the fluid region and $2\,496$ in the solid region. The velocity magnitude is shown in the fluid region while the displacement magnitude is shown in the solid region.}
   \label{fig:hronTurek-results}
\end{figure}
The predicted mean, amplitude and frequency of the vertical and horizontal displacement of the end of the plate are compared with the results from \citet{Turek2006} in Table \ref{tab:hronTurekDisp}.
As directed in \citet{Turek2006}, the maximum and minimum values for the last period are used to calculate the mean and amplitude:
\begin{eqnarray}
	\text{mean} = \frac{1}{2}(\text{max} + \text{min}), \quad\quad
	\text{amplitude} = \frac{1}{2}(\text{max} - \text{min})
\end{eqnarray}
while the frequency is calculated as the inverse of the period.
\begin{table}[htb]
	\centering
		\begin{tabular}{lll}
			\hline
			Mesh (number of fluid and solid cells) & $u_x$ (in mm) & $u_y$ (in mm) \\
			\hline
			%  \hl{This data is taken as the average from 4-6 seconds: I am now running the case up to 20 seconds as per the benchmark}
%			1 ($1\,252$ + 156)  & $-0.1 \pm 0.01 \,[ 18.5 ]$ & $1.60 \pm 0.20 \,[ 5.5 ]$ \\
%			2 ($5\,008$ + 624)  & $-2.13 \pm 2.07 \,[ 11 ]$ & $2.00 \pm 28.84 \,[ 5.5 ]$ \\
%			3 ($20\,032$ + $2\,496$)  & $-2.62 \pm 2.46 \,[11]$ & $1.80 \pm 32.01 \,[ 5.5 ]$ \\
			% Data below is from the last cycle in the case run up to 20 s
			1 ($1\,252$ + 156)  & $ -0.1 \pm 0.001 \,[ 47.6 ]$ & $1.58 \pm 0.002 \,[ 5.88 ]$ \\
			2 ($5\,008$ + 624)  & $-2.16 \pm 2.00 \,[ 10.2 ]$ & $1.54 \pm 29.56 \,[ 5.6 ]$ \\
			3 ($20\,032$ + $2\,496$)  & $-2.51 \pm 2.37 \,[12.1]$ & $0.87 \pm 33.08 \,[ 5.6 ]$ \\
			\hline
			\citet{Turek2006} & $-2.69 \pm 2.53\,[10.9]$ & $1.48 \pm 34.38\,[5.3]$ \\
			\hline
		\end{tabular}
	\caption{Average predicted displacements (mean $\pm$ amplitude $[$frequency$]$) of the free end of the plate. Note that large-scale plate oscillations in mesh 1 (coarsest mesh) die out within the first 10 s.}
%	 \hl{Hron-Turek say to use the last cycle, but surely taking the average of cycles would be better for benchmarking...} \hl{e.g. the average of the mesh 3 cycle is slightly lower than the rest} \hl{we could also report the average}
		\label{tab:hronTurekDisp}
\end{table}

@@

Table \ref{tab:times_memory} lists the wall clock times (time according to a clock on the wall) and maximum memory usage for all cases and all meshes, where diverged cases are indicated by the symbol $\dag$.
Figure \ref{fig:times_memory}(a) plots the corresponding speedup on all cases as a function of the number of degrees of freedom, where the speedup is defined as the segregated clock time divided by the Jacobian-free Newton-Krylov clock time.
The number of degrees of freedom is $2 \times$ the cell count in 2-D cases and $3 \times$ in 3-D cases, except in the fluid-solid interaction case where the fluid domain has three degrees of freedom (two velocity components and pressure) in 2-D.
Speedup is only calculated for clock times greater than 2 seconds to avoid the comparison of small numbers.
In all cases where convergence was achieved, the Jacobian-free Newton-Krylov approach was faster (speedup $> 1$), with speedups of one or two orders of magnitude in many cases.
Additionally, it can be observed from Figure \ref{fig:times_memory}(a) that the speedup increases as the number of degrees of freedom increases.
The largest speedup was 237 on the finest mesh of the membrane i (2-D, linear elastic) case, while the lowest speed was 1.09 and occurred on the coarsest mesh in the fluid-solid interaction case.
Excluding the fluid-solid interaction, where the majority of degrees of freedom are placed in the fluid region, the next lowest speed-up was 1.66 and occurred on a coarse mesh in the membrane ii (2-D, hyperelastic) case.
\begin{table}[!htbp]
	\centering
		\begin{tabular}{ll|ll|ll}
			\hline
			\textbf{Case} & \textbf{Cell Count} & \multicolumn{2}{c|}{\textbf{JFNK}} & \multicolumn{2}{c}{\textbf{Segregated}} \\
			     &            & \textbf{Time} & \textbf{Memory} & \textbf{Time} & \textbf{Memory} \\
			     &            & (in s) & (in MB) & (in s) & (in MB) \\
			\hline
			\input{tableJFNKvsSeg_mms.tex}
			\input{tableJFNKvsSeg_sphericalCavity.tex}
			\input{tableJFNKvsSeg_ellipticPlate.tex}
%			\input{tableJFNKvsSeg_Tmember.tex}
			\input{tableJFNKvsSeg_ventricle.tex}
			\input{tableJFNKvsSeg_membranei.tex}
			\input{tableJFNKvsSeg_membraneii.tex}
			\input{tableJFNKvsSeg_membraneiii.tex}
			\input{tableJFNKvsSeg_cantilever.tex}
%			\input{tableJFNKvsSeg_turbine.tex}
			\input{tableJFNKvsSeg_HronTurek.tex}
		\end{tabular}
	\caption{Execution times (rounded to the nearest second) and maximum memory usage for Jacobian-free Newton-Krylov (JFNK) and segregated methods (rounded to the nearest MB). $\dag$ indicates the solver diverged. }
	\label{tab:times_memory}
\end{table}
\hl{Update results (replace snes.seg with of.seg results}
\hl{Replace ellipticPlate results with sphericalCavity or re-add it}

Figure \ref{fig:speedup_rel_mem}(b) shows the \emph{relative memory} usage, defined here as the maximum memory usage of the segregated approach divided by that of the Jacobian-free Newton-Krylov approach.
The relative memory usage is close to unity in most cases, indicating there is no significant increase in memory requirements when switching from a segregated approach to a Jacobian-free Newton-Krylov approach.
In addition, unlike for the speedup, there is no general trend in the relative memory usage as the number of degrees of freedom increases.
The maximum value for relative memory was 1.9 and occurred in the second-finest mesh for the membrane ii (2-D, hyperelastic) case, although for all other meshes in this case, the value was less than one (between 0.82 and 0.95).
The maximum memory usage is primarily attributed to the linear solver and preconditioner; the choice of preconditioner and linear solver settings are examined further in Section \ref{sec:preconditioner}.
%\hl{Note: we use segregated in PETSc, so it is a vector system}
\begin{figure}[htbp]
   \centering
	\subfigure[Speedup]
	{
   		\includegraphics[width=0.48\textwidth]{figures/speedup}
	}
	\subfigure[Relative memory]
	{
   		\includegraphics[width=0.48\textwidth]{figures/relative_memory}  
   	}
   \caption{The speedup (segregated clock time divided by the Jacobian-free Newton-Krylov clock time) and relative memory usage (segregated maximum memory usage divided by the Jacobian-free Newton-Krylov maximum memory usage) as a function of degrees of freedom}
   \label{fig:speedup_rel_mem}
\end{figure}

>>>>

\author*[1,2,3]{\fnm{Philip} \sur{Cardiff}}\email{philip.cardiff@ucd.ie}
\author[1,2,3]{\fnm{Dylan} \sur{Armfield}}
%\author[5]{\fnm{Hiroaki} \sur{Nishikawa}}
\author[4]{\fnm{\v{Z}eljko} \sur{Tukovi\'{c}}}
\author[4]{\fnm{Ivan} \sur{Batisti\'{c}}}

@@

%\hl{5 - oversolving}
%Comment on over-solving => also applies to the segregated system.
%This could be a parameter we look at.
A final comment on the Jacobian-free Newton-Krylov solution algorithm is the potential importance of \emph{oversolving}.
Here, oversolving refers to solving the linear system to too tight a tolerance during the early Newton iterations, essentially wasting time when the solution is far from the true solution.
In addition, some authors %\cite{See164_and_176_in_Knoll2004}
\cite{Knoll2004} have shown Newton convergence to be worse when the earlier iterations are solved to too tight a tolerance.
The concept of oversolving also applies to segregated solution procedures and has been well-known since the early work of Demird\v{z}i\'{c} and co-workers \cite{Demirdzic1995}, where the residuals are typically reduced by one order of magnitude in the inner linear system.
%The optimal choice of residual reduction for a Jacobian-free Newton-Krylov finite volume solid mechanics procedure is explored in Section \ref{sec:test_cases}. \hl{Check: do we examine this?}
In the current work, the residual is reduced by one order of magnitude in the segregated approach and three orders in the Jacobian-free Newton-Krylov approach.
\hl{Try for EW plastic: $-snes_ksp_ew -snes_ksp_ew_version 2 -snes_ksp_ew_rtol0 1e-4$}

@@

Figure \ref{fig:hronTurek-results} shows the velocity field in the fluid region and displacement magnitude in the solid region at $t = 4.42$ s, corresponding to a peak in the vertical displacement of the plate free-end oscillation.
The fluid is seen to accelerate as it passes the cylinder, with regular vortices being thrown off the plate, causing it to oscillate.
\begin{figure}[htbp]
   \centering
	   \includegraphics[width=\textwidth]{figures/hronTurek-results} 
%	\subfigure[Velocity field in the fluid region and displacement magnitude in the solid region at $t = 4.42$ s]
%	{
%	   \includegraphics[width=0.8\textwidth]{figures/hronTurek-results} 
%   	}
%	\subfigure[Vertical displacement of the end of the plate vs time]
%	{
%   		\includegraphics[width=0.3\textwidth]{figures/placeholder}  
%   	}
   \caption{Elastic plate behind a rigid cylinder case results at $t = 20$ s for the mesh with $20\,032$ cells in the fluid region and $2\,496$ in the solid region. The velocity magnitude is shown in the fluid region while the displacement magnitude is shown in the solid region. \hl{update: harmonic top/bot}}
   \label{fig:hronTurek-results}
\end{figure}
The predicted mean, amplitude and frequency of the vertical and horizontal displacement of the end of the plate are compared with the results from \citet{Turek2006} in Table \ref{tab:hronTurekDisp}.
As directed in \citet{Turek2006}, the maximum and minimum values for the last period are used to calculate the mean and amplitude:
\begin{eqnarray}
	\text{mean} = \frac{1}{2}(\text{max} + \text{min}), \quad\quad
	\text{amplitude} = \frac{1}{2}(\text{max} - \text{min})
\end{eqnarray}
while the frequency is calculated as the inverse of the period.
\begin{table}[htb]
	\centering
		\begin{tabular}{lll}
			\hline
			Mesh (number of fluid and solid cells) & $u_x$ (in mm) & $u_y$ (in mm) \\
			\hline
			1 ($1\,252$ + 156)  & $ -0.1 \pm 0.001 \,[ 47.6 ]$ & $1.58 \pm 0.002 \,[ 5.88 ]$ \\
			2 ($5\,008$ + 624)  & $-2.16 \pm 2.00 \,[ 10.2 ]$ & $1.54 \pm 29.56 \,[ 5.6 ]$ \\
			3 ($20\,032$ + $2\,496$)  & $-2.51 \pm 2.37 \,[12.1]$ & $0.87 \pm 33.08 \,[ 5.6 ]$ \\
			\hline
			\citet{Turek2006} & $-2.69 \pm 2.53\,[10.9]$ & $1.48 \pm 34.38\,[5.3]$ \\
			\hline
		\end{tabular}
	\caption{Average predicted displacements (mean $\pm$ amplitude $[$frequency$]$) of the free end of the plate. Note that large-scale plate oscillations in mesh 1 (coarsest mesh) die out within the first 10 s.}
%	 \hl{Hron-Turek say to use the last cycle, but surely taking the average of cycles would be better for benchmarking...} \hl{e.g. the average of the mesh 3 cycle is slightly lower than the rest} \hl{we could also report the average}
		\label{tab:hronTurekDisp}
\end{table}

@@

Table \ref{tab:times_memory} lists the wall clock times (time according to a clock on the wall) and maximum memory usage for all cases and all meshes, where diverged cases are indicated by the symbol $\dag$.
Figure \ref{fig:times_memory}(a) plots the corresponding speedup on all cases as a function of the number of degrees of freedom, where the speedup is defined as the segregated clock time divided by the Jacobian-free Newton-Krylov clock time.
The number of degrees of freedom is $2 \times$ the cell count in 2-D cases and $3 \times$ in 3-D cases, except in the fluid-solid interaction case where the fluid domain has three degrees of freedom (two velocity components and pressure) in 2-D.
Speedup is only calculated for clock times greater than 2 seconds to avoid the comparison of small numbers.
In all cases where convergence was achieved, the Jacobian-free Newton-Krylov approach was faster (speedup $> 1$), with speedups of one or two orders of magnitude in many cases.
Additionally, it can be observed from Figure \ref{fig:times_memory}(a) that the speedup increases as the number of degrees of freedom increases.
The largest speedup was 237 on the finest mesh of the membrane i (2-D, linear elastic) case, while the lowest speed was 1.09 and occurred on the coarsest mesh in the fluid-solid interaction case.
Excluding the fluid-solid interaction, where the majority of degrees of freedom are placed in the fluid region, the next lowest speed-up was 1.66 and occurred on a coarse mesh in the membrane ii (2-D, hyperelastic) case.
\begin{table}[!htbp]
	\centering
		\begin{tabular}{ll|ll|ll}
			\hline
			\textbf{Case} & \textbf{Cell Count} & \multicolumn{2}{c|}{\textbf{JFNK}} & \multicolumn{2}{c}{\textbf{Segregated}} \\
			     &            & \textbf{Time} & \textbf{Memory} & \textbf{Time} & \textbf{Memory} \\
			     &            & (in s) & (in MB) & (in s) & (in MB) \\
			\hline
			\input{tableJFNKvsSeg_mms.tex}
			\input{tableJFNKvsSeg_sphericalCavity.tex}
			\input{tableJFNKvsSeg_ellipticPlate.tex}
%			\input{tableJFNKvsSeg_Tmember.tex}
			\input{tableJFNKvsSeg_ventricle.tex}
			\input{tableJFNKvsSeg_membranei.tex}
			\input{tableJFNKvsSeg_membraneii.tex}
			\input{tableJFNKvsSeg_membraneiii.tex}
			\input{tableJFNKvsSeg_cantilever.tex}
%			\input{tableJFNKvsSeg_turbine.tex}
			\input{tableJFNKvsSeg_HronTurek.tex}
		\end{tabular}
	\caption{Execution times (rounded to the nearest second) and maximum memory usage for Jacobian-free Newton-Krylov (JFNK) and segregated methods (rounded to the nearest MB). $\dag$ indicates the solver diverged. }
	\label{tab:times_memory}
\end{table}
\hl{Update results (replace snes.seg with of.seg results); running!}

Figure \ref{fig:speedup_rel_mem}(b) shows the \emph{relative memory} usage, defined here as the maximum memory usage of the segregated approach divided by that of the Jacobian-free Newton-Krylov approach.
The relative memory usage is close to unity in most cases, indicating there is no significant increase in memory requirements when switching from a segregated approach to a Jacobian-free Newton-Krylov approach.
In addition, unlike for the speedup, there is no general trend in the relative memory usage as the number of degrees of freedom increases.
The maximum value for relative memory was 1.9 and occurred in the second-finest mesh for the membrane ii (2-D, hyperelastic) case, although for all other meshes in this case, the value was less than one (between 0.82 and 0.95).
The maximum memory usage is primarily attributed to the linear solver and preconditioner; the choice of preconditioner and linear solver settings are examined further in Section \ref{sec:preconditioner}.
\begin{figure}[htbp]
   \centering
	\subfigure[Speedup]
	{
   		\includegraphics[width=0.48\textwidth]{figures/speedup}
	}
	\subfigure[Relative memory]
	{
   		\includegraphics[width=0.48\textwidth]{figures/relative_memory}  
   	}
   \caption{The speedup (segregated clock time divided by the Jacobian-free Newton-Krylov clock time) and relative memory usage (segregated maximum memory usage divided by the Jacobian-free Newton-Krylov maximum memory usage) as a function of degrees of freedom}
   \label{fig:speedup_rel_mem}
\end{figure}
